---
title: "Code_For_KC_MS"
author: "Victoria Fisher"
date: "2025-01-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Libraries and Data

```{r}
library(pacman)
p_load(tidyverse,
       tidycensus,
       ggplot2,
       gtsummary,
       DescTools,
       datawizard,
       performance,
       stats,
       glmmTMB,
       zctaCrosswalk)

#Mortality Data

load('[project folder here]/all_deaths_2015.rda')
load('[project folder here]/all_deaths_2016.rda')
load('[project folder here]/all_deaths_2017.rda')
load('[project folder here]/all_deaths_2018.rda')
load('[project folder here]/all_deaths_2019.rda')
load('[project folder here]/all_deaths_2020.rda')
load('[project folder here]/all_deaths_2021.rda')


#NaNDA Data

load("[project folder here]/nanda_socioecon_2020.rda")

socio_econ <- da38528.0005 #rename

#Crosswalk data

load("[project folder here]/ZCTA_county.rda")

load("[project folder here]/county_state.rda")

```

#Pre-COVID Mortality

```{r}

all_deaths_2015$Year <- "2015"
all_deaths_2016$Year <- "2016"
all_deaths_2017$Year <- "2017"
all_deaths_2018$Year <- "2018"
all_deaths_2019$Year <- "2019"

pre_covid_mort <- rbind(all_deaths_2015, all_deaths_2016, all_deaths_2017, all_deaths_2018, all_deaths_2019)

pre_covid_mort <- separate_wider_delim(pre_covid_mort, col = 2, delim = ", ", names = c("County", "State"))

pre_covid_mort$County <- as.factor(pre_covid_mort$County)

pre_covid_mort$State <- as.factor(pre_covid_mort$State)

#Filtering to Kansas City Metro Area Counties
pre_covid_mort <- filter(pre_covid_mort, County == "Jackson County" & State == "MO" | County == "Johnson County" & State == "KS" | County == "Clay County" & State == "MO" | County == "Wyandotte County" | County == "Cass County" | County == "Platte County" | County == "Miami County" | County ==  "Leavenworth County" | County == "Lafayette County" | County == "Ray County" | County == "Clinton County" | County == "Bates County" | County == "Linn County" & State == "KS" | County == "Caldwell County")

pre_covid_mort <- droplevels(pre_covid_mort)

colnames(pre_covid_mort)[7] <- "crude_rate"
colnames(pre_covid_mort)[4] <- "GEOID"

pre_covid_mort$crude_rate <- as.numeric(pre_covid_mort$crude_rate)

#Adding older age

#Adding predictors of mortality
v2015 <- load_variables(2015, "acs5")

vars <- c(total_pop = "B01003_001", 
          male_70_74 = "B01001_022",
          male_75_79 = "B01001_023",
          male_80_84 = "B01001_024",
          male_85_plus = "B01001_025",
          female_70_74 = "B01001_046",
          female_75_79 = "B01001_047",
          female_80_84 = "B01001_048",
          female_85_plus = "B01001_049"
          )

#Pulling age
years <- 2015:2019 #pulls 5-year estimate American Community Survey data from 2015 - 2019
names(years) <- years
mort_vars <- purrr::map_dfr(years, ~{
  get_acs(
    geography = "County", 
    variables = vars, 
    geometry = FALSE,
    year = .x
    )
}, .id = "year")

mort_vars <- mort_vars %>%
  pivot_wider(id_cols = c("year", "GEOID"),
              names_from = c("variable"),
              values_from = "estimate")

mort_vars$older_age <- rowSums(mort_vars[, 3:10])

mort_vars <- mort_vars %>% dplyr::select(!c(3:10))

mort_vars$P_older_age <- (mort_vars$older_age / mort_vars$total_pop) * 100

colnames(pre_covid_mort)[4] <- "GEOID"

pre_covid_mort <- merge(pre_covid_mort, mort_vars, by.x = c("GEOID", "Year") , by.y = c("GEOID", "year"))

```

#COVID_era mortality

```{r}
#Observed data

#2020 all deaths

all_deaths_2020 <- separate_wider_delim(all_deaths_2020, col = 2, delim = ", ", names = c("County", "State"))

all_deaths_2020 <- filter(all_deaths_2020, County == "Jackson County" & State == "MO" | County == "Johnson County" & State == "KS" | County == "Clay County" & State == "MO" | County == "Wyandotte County" | County == "Cass County" | County == "Platte County" | County == "Miami County" | County ==  "Leavenworth County" | County == "Lafayette County" | County == "Ray County" | County == "Clinton County" | County == "Bates County" | County == "Linn County" & State == "KS" | County == "Caldwell County")

colnames(all_deaths_2020)[7] <- "crude_rate"

all_deaths_2020$Year <- "2020"
colnames(all_deaths_2020)[4] <- "GEOID"


#2021 all deaths

all_deaths_2021 <- separate_wider_delim(all_deaths_2021, col = 2, delim = ", ", names = c("County", "State"))

all_deaths_2021 <- filter(all_deaths_2021, County == "Jackson County" & State == "MO" | County == "Johnson County" & State == "KS" | County == "Clay County" & State == "MO" | County == "Wyandotte County" | County == "Cass County" | County == "Platte County" | County == "Miami County" | County ==  "Leavenworth County" | County == "Lafayette County" | County == "Ray County" | County == "Clinton County" | County == "Bates County" | County == "Linn County" & State == "KS" | County == "Caldwell County")

colnames(all_deaths_2021)[7] <- "crude_rate"

all_deaths_2021$Year <- "2021"

colnames(all_deaths_2021)[4] <- "GEOID"


#Adding COVID-era older age (mortality adjustment)
v2020 <- load_variables(2020, "acs5")

vars <- c(total_pop = "B01003_001", 
          male_70_74 = "B01001_022",
          male_75_79 = "B01001_023",
          male_80_84 = "B01001_024",
          male_85_plus = "B01001_025",
          female_70_74 = "B01001_046",
          female_75_79 = "B01001_047",
          female_80_84 = "B01001_048",
          female_85_plus = "B01001_049"
          )
#Adding drove alone
years <- 2020:2021 #pulls American Community Survey data from 2020 - 2021
names(years) <- years
mort_vars_20_21 <- purrr::map_dfr(years, ~{
  get_acs(
    geography = "County", 
    variables = vars, 
    geometry = FALSE,
    year = .x
    )
}, .id = "year")

mort_vars_20_21 <- mort_vars_20_21 %>%
  pivot_wider(id_cols = c("year", "GEOID"),
              names_from = c("variable"),
              values_from = "estimate")

mort_vars_20_21$older_age <- rowSums(mort_vars_20_21[, 3:10])

mort_vars_20_21 <- mort_vars_20_21 %>% dplyr::select(!c(3:10))

mort_vars_20_21$P_older_age <- (mort_vars_20_21$older_age / mort_vars_20_21$total_pop) * 100

older_age_20 <- filter(mort_vars_20_21, year == 2020)

all_deaths_2020 <- merge(all_deaths_2020, older_age_20, by.x = c("GEOID", "Year") , by.y = c("GEOID", "year"))

older_age_21 <- filter(mort_vars_20_21, year == 2021)

all_deaths_2021 <- merge(all_deaths_2021, older_age_21, by.x = c("GEOID", "Year") , by.y = c("GEOID", "year"))


```

#Regression for predicting expected deaths in 2020 and 2021

```{r}
#Deaths predicted 2020

glm_fit <- glmmTMB(Deaths ~ P_older_age + offset(log(Population)) + (1 | Year) + (1 | County), family = "nbinom1", pre_covid_mort)
pred_df <- data.frame(Deaths = predict(glm_fit, pre_covid_mort, type = "response"), County = pre_covid_mort$County, Year = pre_covid_mort$Year)

#predicted raw deaths 2020

train_2020 <- all_deaths_2020 %>% dplyr::select(!c("Deaths"))

predict.m1b.2020 <- data.frame(exp_deaths = predict(glm_fit, newdata = train_2020, type = "response", allow.new.levels = TRUE), GEOID = unique(pre_covid_mort$GEOID), Year = "2020")

#predicted raw deaths 2021
train_2021 <- all_deaths_2021 %>% dplyr::select(!c("Deaths"))

predict.m1b.2021 <- data.frame(exp_deaths = predict(glm_fit, newdata = train_2021, type = "response", allow.new.levels = TRUE), GEOID = unique(pre_covid_mort$GEOID), Year = "2021")


#Excess mortality

df_exp_deaths <- rbind(predict.m1b.2020, predict.m1b.2021)

df_obs_deaths <- rbind(all_deaths_2020, all_deaths_2021)

df_excess_mort <- merge(df_obs_deaths, df_exp_deaths, by = c("Year", "GEOID"))

df_excess_mort$excess_mortality <- df_excess_mort$Deaths - df_excess_mort$exp_deaths

```

#ZCTA and County Crosswalks

```{r}

#Edit County-State dataframe

county_state <- separate_wider_delim(county_state, col = 1, delim = ", ", names = c("County", "State"))

county_state$County <- as.factor(county_state$County)

county_state$State <- as.factor(county_state$State)

colnames(county_state)[3] <- "GEOID_COUNTY_20"

county_state$GEOID_COUNTY_20 <- as.factor(county_state$GEOID_COUNTY_20)

county_state <- county_state %>%dplyr::select(c(1:3))

county_state <- county_state %>% distinct(GEOID_COUNTY_20, State, County)

#Merge zcta and county

county_state_zcta <- merge(ZCTA_County, county_state, by = "GEOID_COUNTY_20")

#Merge with socioecon

socio_zcta_county_state <- merge(county_state_zcta, socio_econ, by = "ZCTA20")

#Reduce to KCMA

states <- c("Kasnas", "Missouri")

ks_zcta <- as.data.frame(get_zctas_by_state("Kansas"))
mo_zcta <- as.data.frame(get_zctas_by_state("Missouri"))

counties <- c("Jackson County", "Johnson County", "Clay County", "Wyandotte County", "Cass County","Platte County", "Miami County", "Leavenworth County", "Lafayette County", "Ray County", "Clinton County", "Bates County", "Linn County", "Caldwell County")

colnames(ks_zcta)[1] <- "ZCTA20"
colnames(mo_zcta)[1] <- "ZCTA20"

ks_zcta$State <- "KS"
mo_zcta$State <- "MO"

ks_mo_ZCTA <- rbind(ks_zcta, mo_zcta)

ks_mo_ZCTA$ZCTA20 <- as.factor(ks_mo_ZCTA$ZCTA20)
ks_mo_ZCTA$State <- as.factor(ks_mo_ZCTA$State)

socio_ks_mo <- merge(socio_zcta_county_state, ks_mo_ZCTA, by = c("ZCTA20", "State"))

colnames(socio_ks_mo)[12] <- "County_fac"

socio_ks_mo$County_fac <- as.factor(socio_ks_mo$County_fac)

#ZCTA level dataset

df_kcma_zcta <- filter(socio_ks_mo, County_fac == "Jackson County" & State == "MO" | County_fac == "Johnson County" & State == "KS" | County_fac == "Clay County" & State == "MO" | County_fac == "Wyandotte County" & State == "KS" | County_fac == "Cass County" & State == "MO" | County_fac == "Platte County" | County_fac == "Miami County" & State == "KS" | County_fac ==  "Leavenworth County" & State == "KS" | County_fac == "Lafayette County"  | County_fac == "Ray County" & State == "MO"  | County_fac == "Clinton County" | County_fac == "Bates County" & State == "MO" | County_fac == "Linn County" & State == "KS" | County_fac == "Caldwell County" & State == "MO" )

```

#Excess mortality dataset

```{r}

df_excess_mort$County <- as.factor(df_excess_mort$County)
df_excess_mort$State <- as.factor(df_excess_mort$State)

colnames(df_kcma_zcta)[3] <- "GEOID"

df_kcma_zcta_mort <- merge(df_excess_mort, df_kcma_zcta, by = c("GEOID", "County", "State"))

```

#ASC Variables

```{r}

df_zcta <- df_kcma_zcta_mort

v2020 <- load_variables(2020, "acs5")

vars <- c(zcta_pop = "B01003_001",
          zcta_male_pop = "B01001_002", 
          zcta_work_from_home = "B08006_017",
          zcta_tot_pub_assist = "B09010_002",
          zcta_medincome = "B07011_001",
          zcta_medicaid_under19 = "B27010_006",
          zcta_medicaid_19_34 = "B27010_022",
          zcta_medicaid_35_64 = "B27010_038",
          zcta_medicaid_over64 = "B27010_055",
          zcta_tenure_total = "B25003_001",
          zcta_owner_occ = "B25003_002",
          zcta_did_not_wfh = "B99084_002"
          )
#Adding drove alone
years <- 2020:2021 #pulls American Community Survey data from 2019 - 2021
names(years) <- years
zcta_acs_vars <- purrr::map_dfr(years, ~{
  get_acs(
    geography = "zcta", 
    variables = vars, 
    geometry = TRUE,
    year = .x
    )
}, .id = "year")

colnames(zcta_acs_vars)[2] <- "ZCTA20"
colnames(zcta_acs_vars)[1] <- "Year"

zcta_acs_vars <- zcta_acs_vars %>%
  pivot_wider(id_cols = c("Year", "ZCTA20", "geometry"),
              names_from = c("variable"),
              values_from = "estimate")


zcta_acs_vars$ZCTA20 <- as.factor(zcta_acs_vars$ZCTA20)
zcta_acs_vars$Year <- as.factor(zcta_acs_vars$Year)

zcta_acs_vars$medicaid_only <- zcta_acs_vars$zcta_medicaid_under19 + zcta_acs_vars$zcta_medicaid_19_34 + zcta_acs_vars$zcta_medicaid_35_64 + zcta_acs_vars$zcta_medicaid_over64

zcta_acs_vars <- zcta_acs_vars %>% dplyr::select(!c("zcta_medicaid_under19", "zcta_medicaid_19_34", "zcta_medicaid_35_64", "zcta_medicaid_over64"))

df_zcta$Year <- as.factor(df_zcta$Year)

df_zcta <- left_join(df_zcta, zcta_acs_vars, by = c("ZCTA20", "Year"))
summary(is.na(df_zcta))

#Removing variables with high missingness 
df_zcta <- df_zcta %>% select(!c("MEDFAMINC_NHWHITE16_20", "PFAMINCGE40LT75K16_20", "PFAMINCGE75LT125K16_20", "PFAMINCLT40K16_20", "PFAMINCGE125K16_20", "MEDFAMINC_BLACK16_20", "RATIO_MEDFAMINC_NHWTOHISP16_20", "RATIO_MEDFAMINC_NHWTOB16_20", "MEDFAMINC_HISPANIC16_20", "PSNGPNT16_20", "PFHFAM16_20"))

df_zcta <- df_zcta %>% relocate(c(zcta_medincome, zcta_owner_occ, zcta_tenure_total), .before = zcta_male_pop)

percent_cols <- names(df_zcta[, 63:67])

df_zcta <- df_zcta %>%
  mutate(across(all_of(percent_cols), ~ (. / zcta_pop) * 100, .names = "P_{.col}"))

df_zcta <- df_zcta %>%
  mutate(across(ends_with("16_20"), ~ .x * 100, .names = "P_{.col}"))

df_zcta$P_owner_occ <- (df_zcta$zcta_owner_occ / df_zcta$zcta_tenure_total) * 100

df_zcta$P_not_owner_occ <- 100 - df_zcta$P_owner_occ

df_zcta$P_no_bach <- 100 - df_zcta$P_PED3_16_20


#Removing rows with high missingness 

df_zcta.na <- df_zcta %>% filter(!is.na(MEDFAMINC16_20))

df_zcta.na <- df_zcta.na %>% filter(!is.na(zcta_medincome))

df_zcta.na$stnd_medincome <- standardise(df_zcta.na$zcta_medincome)

```

#Removing duplicates

```{r}
length(unique(df_zcta.na$ZCTA20))

df_zcta.na$zcta_num <- as.numeric((as.character(df_zcta.na$ZCTA20)))

df_2020 <- filter(df_zcta.na, Year == "2020")

df_dupe <- df_2020 %>% dplyr::distinct(zcta_num, County)

df_dupe <- df_dupe %>%
   group_by(zcta_num) %>%
   filter(n()>1)

length(unique(df_dupe$zcta_num))
length(unique(df_zcta.na$zcta_num))

df_dedupe <- filter(df_dupe, zcta_num == "64017" & County == "Ray County" | zcta_num == "64024" & County == "Clay County" | zcta_num == "64034" & County == "Jackson County" | zcta_num == "64048" & County == "Clay County" |  zcta_num == "64061" & County == "Johnson County" | zcta_num == "64062" & County == "Ray County" | zcta_num == "64075" & County == "Jackson County" | zcta_num == "64077" & County == "Ray County" | zcta_num == "64080" & County == "Cass County" | zcta_num == "64082" & County == "Jackson County" | zcta_num == "64089" & County == "Clay County" | zcta_num == "64118" & County == "Clay County" |  zcta_num == "64147" & County == "Jackson County" | zcta_num == "64150" & County == "Platte County" | zcta_num == "64151" & County == "Platte County" | zcta_num == "64155" & County == "Clay County" | zcta_num == "64429" & County == "Clinton County" | zcta_num == "64444" & County == "Platte County" | zcta_num == "64465" & County == "Clinton County" | zcta_num == "64492" & County == "Clinton County" | zcta_num == "64624" & County == "Caldwell County" | zcta_num == "64637" & County == "Caldwell County" | zcta_num == "64671" & County == "Caldwell County" | zcta_num == "64725" & County == "Cass County" | zcta_num == "64742" & County == "Cass County" | zcta_num == "66012" & County == "Wyandotte County" | zcta_num == "66013" & County == "Johnson County" | zcta_num == "66021" & County == "Johnson County" | zcta_num == "66026" & County == "Miami County" | zcta_num == "66040" & County == "Linn County" | zcta_num == "66072" & County == "Linn County" | zcta_num == "66083" & County == "Johnson County" | zcta_num == "66109" & County == "Wyandotte County" | zcta_num == "66216" & County == "Johnson County")

to_remove <- filter(df_dupe, !(zcta_num == "64017" & County == "Ray County" | zcta_num == "64024" & County == "Clay County" | zcta_num == "64034" & County == "Jackson County" | zcta_num == "64048" & County == "Clay County" |  zcta_num == "64061" & County == "Johnson County" | zcta_num == "64062" & County == "Ray County" | zcta_num == "64075" & County == "Jackson County" | zcta_num == "64077" & County == "Ray County" | zcta_num == "64080" & County == "Cass County" | zcta_num == "64082" & County == "Jackson County" | zcta_num == "64089" & County == "Clay County" | zcta_num == "64118" & County == "Clay County" |  zcta_num == "64147" & County == "Jackson County" | zcta_num == "64150" & County == "Platte County" | zcta_num == "64151" & County == "Platte County" | zcta_num == "64155" & County == "Clay County" | zcta_num == "64429" & County == "Clinton County" | zcta_num == "64444" & County == "Platte County" | zcta_num == "64465" & County == "Clinton County" | zcta_num == "64492" & County == "Clinton County" | zcta_num == "64624" & County == "Caldwell County" | zcta_num == "64637" & County == "Caldwell County" | zcta_num == "64671" & County == "Caldwell County" | zcta_num == "64725" & County == "Cass County" | zcta_num == "64742" & County == "Cass County" | zcta_num == "66012" & County == "Wyandotte County" | zcta_num == "66013" & County == "Johnson County" | zcta_num == "66021" & County == "Johnson County" | zcta_num == "66026" & County == "Miami County" | zcta_num == "66040" & County == "Linn County" | zcta_num == "66072" & County == "Linn County" | zcta_num == "66083" & County == "Johnson County" | zcta_num == "66109" & County == "Wyandotte County" | zcta_num == "66216" & County == "Johnson County"))

length(unique(to_remove$zcta_num))

df_zcta.na <- anti_join(df_zcta.na, to_remove, by = c("zcta_num", "County"))

length(unique(df_zcta.na$ZCTA20))

table(df_zcta.na$Year)

table(df_zcta.na$ZCTA20)

df_zcta.na <- filter(df_zcta.na, !(ZCTA20 == "64149" | ZCTA20 == "64066" | ZCTA20 == "64167"))

table(df_zcta.na$Year) #210 both years


```

#Estimating excess mortality at the ZIP level

```{r}
#ZIP-code risk score

df_zcta.na$mort_rate_100k <- (df_zcta.na$Deaths / df_zcta.na$Population) * 100000
df_zcta.na$expected_mort_rate_100k <- (df_zcta.na$exp_deaths / df_zcta.na$Population) * 100000
hist(df_zcta.na$mort_rate_100k)

county_summary <- df_zcta.na %>%
  group_by(County, Year) %>%
  summarise_at(c("Deaths", "exp_deaths"), ~mean(.x))

#Weights

glm_weights <- glmmTMB(Deaths ~ P_PGE70_16_20 + offset(log(zcta_pop)), family = "nbinom1", df_zcta.na) #Model is the same as the county predicted deaths
summary(glm_weights)
check_model(glm_weights)

df_zcta.na$risk_score <- (df_zcta.na$P_PGE70_16_20 * 0.05894)
summary(df_zcta.na$risk_score)

#df_zcta.na$risk_score <- ifelse(df_zcta.na$risk_score <=0, 0.01, df_zcta.na$risk_score)

df_zcta.na <- df_zcta.na %>%
  group_by(County, Year) %>%
  mutate(
    normalized_weight = risk_score / sum(risk_score),  # Normalize risk weights
    estimated_zcta_deaths = normalized_weight * unique(Deaths), #allocate observed deaths
    expected_zcta_deaths = normalized_weight * unique(exp_deaths)# Allocate expected deaths
  )

by(df_zcta.na$estimated_zcta_deaths, df_zcta.na$Year, summary)

by(df_zcta.na$expected_zcta_deaths, df_zcta.na$Year, summary)

# Sum of estimated ZCTA-level deaths within each county

zcta_summary <- df_zcta.na %>%
  group_by(County, Year) %>%
  summarise_at("estimated_zcta_deaths", ~sum(.x))

county_summary <- df_zcta.na %>%
  group_by(County, Year) %>%
  summarise_at("Deaths", ~mean(.x))

#Creating estimated ZCTA death counts

df_zcta.na$estimated_zcta_death_rate <- (df_zcta.na$estimated_zcta_deaths / df_zcta.na$zcta_pop) * 10000
df_zcta.na$expected_zcta_death_rate <- (df_zcta.na$expected_zcta_deaths / df_zcta.na$zcta_pop) * 10000


# Check differences

df_zcta.na$zcta_excess_mort <- df_zcta.na$estimated_zcta_deaths - df_zcta.na$expected_zcta_deaths

#Dealing with negative  - because we are looking at "excess" mortality, we replace negative predicted mortality numbers with 0. It is possible to have negative excess mortality, especially if these ZCTAs/counties were especiially adherent to public health mandates and measures in preventing spread during the COVID-19 pandemic. This, however, presents a different methodological questions and challenge and is beyond the scope of the current analysis.

df_excess_mort_less_than_0 <- filter(df_zcta.na, zcta_excess_mort < 0)

df_zcta.na$zcta_excess_mort <- ifelse(df_zcta.na$zcta_excess_mort < 0, 0, df_zcta.na$zcta_excess_mort)

#Mortality rate

df_zcta.na$zcta_excess_rate_10k <- (df_zcta.na$zcta_excess_mort / df_zcta.na$zcta_pop) * 10000

```

# index with scales individually

```{r}
#Obs ZCTA

#Contemporary index 

df_zcta.na$contemp_index <- as.numeric(standardise(df_zcta.na$P_PPUBAS16_20) + standardise(df_zcta.na$P_PUNEMP16_20) + standardise(df_zcta.na$P_medicaid_only) + standardise(df_zcta.na$P_zcta_did_not_wfh))


summary(df_zcta.na$contemp_index)
hist(df_zcta.na$contemp_index)

df_zcta.na$contemp_index_stnd <- standardise(df_zcta.na$contemp_index)
hist(df_zcta.na$contemp_index_stnd)
summary(df_zcta.na$contemp_index_stnd)

df_zcta.na$contemp_index_stnd <- df_zcta.na$contemp_index_stnd + -1 * min(df_zcta.na$contemp_index_stnd)
summary(df_zcta.na$contemp_index_stnd)


#Historical index 

df_zcta.na$hist_index <- as.numeric(standardise(df_zcta.na$P_no_bach) + standardise(df_zcta.na$POPDEN16_20)  + standardise(df_zcta.na$P_not_owner_occ) - standardise(df_zcta.na$P_P18_29_16_20)) 
summary(df_zcta.na$hist_index)
hist(df_zcta.na$hist_index)

df_zcta.na$hist_index_stnd <- standardise(df_zcta.na$hist_index)
summary(df_zcta.na$hist_index_stnd)
df_zcta.na$hist_index_stnd <- df_zcta.na$hist_index_stnd + -1 * min(df_zcta.na$hist_index_stnd)
summary(df_zcta.na$hist_index_stnd)


#Combined index

df_zcta.na$comb_ind <- df_zcta.na$hist_index_stnd + df_zcta.na$contemp_index_stnd
class(df_zcta.na$comb_ind)
hist(df_zcta.na$comb_ind)
summary(df_zcta.na$comb_ind)


df_zcta.na$comb_index_stnd <- standardise(df_zcta.na$hist_index_stnd + df_zcta.na$contemp_index_stnd)
summary(df_zcta.na$comb_index_stnd)
hist(df_zcta.na$comb_index_stnd)

df_zcta.na$comb_index_stnd <- df_zcta.na$comb_index_stnd + -1 * min(df_zcta.na$comb_index_stnd)
summary(df_zcta.na$comb_index_stnd)

```

#Dichotomous indices

```{r}

#Historical dichotomous index
df_zcta.na$hist_dich <- ifelse(df_zcta.na$hist_index_stnd <= mean(df_zcta.na$hist_index_stnd), 0, 1)
df_zcta.na$hist_dich <- as.factor(df_zcta.na$hist_dich)

#Contemporary dichotomous index
summary(df_zcta.na$contemp_index_stnd)
df_zcta.na$contemp_dich <- ifelse(df_zcta.na$contemp_index_stnd <= mean(df_zcta.na$contemp_index_stnd), 0, 1)
df_zcta.na$contemp_dich <- as.factor(df_zcta.na$contemp_dich)

#Combined dichotomous index
summary(df_zcta.na$comb_index_stnd)
df_zcta.na$comb_dich <- ifelse(df_zcta.na$comb_index_stnd <= mean(df_zcta.na$comb_index_stnd), 0, 1)
df_zcta.na$comb_dich <- as.factor(df_zcta.na$comb_dich)

#Combined quadrant index 

df_zcta.na <- df_zcta.na %>% mutate(
  comb_quad = case_when(
    contemp_dich == "0" & hist_dich == "0" ~ "neither",
    contemp_dich == "0" & hist_dich == "1" ~ "above avg hist disen only",
    contemp_dich == "1" & hist_dich == "0" ~ "above avg cont disad only",
    contemp_dich == "1" & hist_dich == "1"~ "above avg in both"
))

df_zcta.na$comb_quad <- as.factor(df_zcta.na$comb_quad)

```

#Descriptive statistics for each variable

```{r}
Desc(df_zcta.na)
```

#Models

```{r}
#Undajusted

glm0 <- glmmTMB(zcta_excess_mort ~ 1 + offset(log(zcta_pop)), family = tweedie(link = "log"),
                data = df_zcta.na)
AIC(glm0)
check_model(glm0) 
qqnorm(resid(glm0))

#Binary indices w state and year

glm2 <- glmmTMB(zcta_excess_mort ~ Year + State + offset(log(zcta_pop)), family = tweedie(link = "log"),
                data = df_zcta.na)
summary(glm2)
confint(glm2)
AIC(glm2)
glm2 %>% tbl_regression(exponentiate = TRUE)
check_model(glm2)

glm3 <- glmmTMB(zcta_excess_mort ~ hist_dich + Year + State + offset(log(zcta_pop)), family = tweedie(link = "log"),
                data = df_zcta.na)
summary(glm3)
confint(glm3)
AIC(glm3)
glm3 %>% tbl_regression(exponentiate = TRUE)
check_model(glm3)

glm4 <- glmmTMB(zcta_excess_mort ~ contemp_dich + State + Year + offset(log(zcta_pop)), family = tweedie(link = "log"),
                data = df_zcta.na)
summary(glm4)
confint(glm4)
AIC(glm4)
glm4 %>% tbl_regression(exponentiate = TRUE)

glm5 <- glmmTMB(zcta_excess_mort ~ comb_dich + State + Year + offset(log(zcta_pop)), family = tweedie(link = "log"),
                data = df_zcta.na)
summary(glm5)
confint(glm5)
AIC(glm5)
glm5 %>% tbl_regression(exponentiate = TRUE)
check_model(glm5)


glm6 <- glmmTMB(zcta_excess_mort ~ relevel(comb_quad, ref = "neither") + State + Year + offset(log(zcta_pop)),  family = tweedie(link = "log"),
                data = df_zcta.na)
summary(glm6)
confint(glm6)
AIC(glm6)
glm6 %>% tbl_regression(exponentiate = TRUE)
check_model(glm6)

```

#combined standard indices, unadjusted

```{r}

glm7 <- glmmTMB(zcta_excess_mort ~ hist_index_stnd + offset(log(zcta_pop)),  family = tweedie(link = "log"),
                data = df_zcta.na)
summary(glm7)
glm7 %>% tbl_regression(exponentiate = TRUE)


glm8 <- glmmTMB(zcta_excess_mort ~ contemp_index_stnd +  offset(log(zcta_pop)),  family = tweedie(link = "log"),
                data = df_zcta.na)
summary(glm8)
AIC(glm8)
glm8 %>% tbl_regression(exponentiate = TRUE)

glm9 <- glmmTMB(zcta_excess_mort ~ comb_index_stnd + offset(log(zcta_pop)),  family = tweedie(link = "log"),
                data = df_zcta.na)
summary(glm9)
BIC(glm9)
glm9 %>% tbl_regression(exponentiate = TRUE)
check_model(glm9)
```


#Sensitivity Analyses with race

```{r}
#Tests for significance

t.test(df_zcta.na$P_PNHBLACK16_20 ~ df_zcta.na$hist_dich)
t.test(df_zcta.na$P_PNHBLACK16_20 ~ df_zcta.na$contemp_dich)

#Contemporary index with race

df_zcta.na$contemp_index_w_race <- as.numeric(standardise(df_zcta.na$P_PPUBAS16_20) + standardise(df_zcta.na$P_PUNEMP16_20) + standardise(df_zcta.na$P_medicaid_only) + standardise(df_zcta.na$P_zcta_did_not_wfh) + standardise(df_zcta.na$P_PNHBLACK16_20))

summary(df_zcta.na$contemp_index_w_race)
hist(df_zcta.na$contemp_index)

df_zcta.na$contemp_index_w_race_stnd <- standardise(df_zcta.na$contemp_index_w_race)
hist(df_zcta.na$contemp_index_w_race_stnd)
summary(df_zcta.na$contemp_index_w_race_stnd)

df_zcta.na$contemp_index_w_race_stnd <- df_zcta.na$contemp_index_w_race_stnd + -1 * min(df_zcta.na$contemp_index_w_race_stnd)
summary(df_zcta.na$contemp_index_stnd)

t.test(df_zcta.na$contemp_index_w_race_stnd, df_zcta.na$contemp_index_stnd)
cor.test(df_zcta.na$contemp_index_w_race_stnd, df_zcta.na$contemp_index_stnd) #.93

#Historical index with race

df_zcta.na$hist_index_w_race <- as.numeric(standardise(df_zcta.na$P_no_bach) + standardise(df_zcta.na$POPDEN16_20)  + standardise(df_zcta.na$P_not_owner_occ) - standardise(df_zcta.na$P_P18_29_16_20) + standardise(df_zcta.na$P_PNHBLACK16_20)) 
summary(df_zcta.na$hist_index_w_race)
hist(df_zcta.na$hist_index_w_race)

df_zcta.na$hist_index_w_race_stnd <- standardise(df_zcta.na$hist_index_w_race)
summary(df_zcta.na$hist_index_w_race_stnd)
df_zcta.na$hist_index_w_race_stnd <- df_zcta.na$hist_index_w_race_stnd + -1 * min(df_zcta.na$hist_index_w_race_stnd)
summary(df_zcta.na$hist_index_w_race_stnd)

cor.test(df_zcta.na$hist_index_w_race_stnd, df_zcta.na$hist_index_stnd) #.92


#Combined index with race

df_zcta.na$comb_ind_w_race <- df_zcta.na$hist_index_w_race_stnd + df_zcta.na$contemp_index_w_race_stnd
class(df_zcta.na$comb_ind_w_race)
hist(df_zcta.na$comb_ind_w_race)
summary(df_zcta.na$comb_ind_w_race)


df_zcta.na$comb_ind_w_race_stnd <- standardise(df_zcta.na$hist_index_w_race_stnd + df_zcta.na$contemp_index_w_race_stnd)
summary(df_zcta.na$comb_ind_w_race_stnd)
hist(df_zcta.na$comb_ind_w_race_stnd)

df_zcta.na$comb_ind_w_race_stnd <- df_zcta.na$comb_ind_w_race_stnd + -1 * min(df_zcta.na$comb_ind_w_race_stnd)
summary(df_zcta.na$comb_ind_w_race_stnd)

cor.test(df_zcta.na$comb_ind_w_race_stnd, df_zcta.na$contemp_index_stnd) #.79

```

